---
title: "BIMM 143 Lab 7: Analyzing sequencing data in the cloud"
author: "Sofia Lanaspa, A17105313"
format: html
---

Successfully initiated launch of instance (i-02966b1f383463c8c)

ssh -i "~/Downloads/sofia_bimm143.pem" ubuntu@ec2-52-12-164-59.us-west-2.compute.amazonaws.com


cd ~/Downloads used to go from home to downloads
if I just put cd Downloads then it tried to get there from the place I am in right now

gunzip: unzip files
tar -xvf: untar files
cd: change dir

PATH is a special evironment variable that store the location of exe files in UNIX systems. I can print it with 'echo $PATH'

We can add to it with:
export PATH=$PATH:/my/new/path

The SRA-TOOLKIT allows us to query and download raw sequencing data from NBCI's SRA database. Key commands include:

prefetch [SRACODE]
fastq-dump [SRACODE]


to connect: 
MacBook-Pro-4:BIMM 143 Lab 17 sofialanaspa$ ls
BIMM 143 Lab 17.Rproj   BIMM143_Lab17.qmd
MacBook-Pro-4:BIMM 143 Lab 17 sofialanaspa$ cd ~/Downloads
MacBook-Pro-4:Downloads sofialanaspa$ chmod 400 "sofia_bimm143.pem"
MacBook-Pro-4:Downloads sofialanaspa$ ssh -i "sofia_bimm143.pem" ubuntu@ec2-52-38-137-79.us-west-2.compute.amazonaws.com


TO SPEED UP KALLISTO:
index so it can go though sequences instead of having to ream from the beginning

$ kallisto quant -i hg19.ensembl -o SRR2156848_quant SRR2156848_1.fastq SRR2156848_2.fastq

$ kallisto quant -i hg19.ensembl -o SRR2156849_quant SRR2156849_1.fastq SRR2156849_2.fastq &

$ kallisto quant -i hg19.ensembl -o SRR2156850_quant SRR2156850_1.fastq SRR2156850_2.fastq &

$ kallisto quant -i hg19.ensembl -o SRR2156851_quant SRR2156851_1.fastq SRR2156851_2.fastq &


chmod +x run.sh
nano
nano run.sh

kallisto quant -i hg19.ensembl -o SRR2156848_quant SRR2156848_1.fastq SRR2156848_2.fastq



scp: secure copy from or to a remote machine
# scp -i keyfile username@remotecomputer:FILE2COPY NEWlocalFILE

scp -r -i "~/Downloads/sofia_bimm143.pem" ubuntu@ec2-52-12-164-59.us-west-2.compute.amazonaws.com:~/*_quant .

-r tells it to go into subdirectories


key class 17: 
ssh -i "sofia_bimm143.pem" ubuntu@ec2-50-112-49-91.us-west-2.compute.amazonaws.com

## Downstream Analysis
```{r}
# BiocManager::install("tximport") 
```

```{r}
library(tximport)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

