---
title: "BIMM 143 Lab 15: RNA-Seq analysis mini-project"
author: Sofia Lanaspa, A17105313
format: pdf
---
Differential expression analysis project with DESeq2 followed by gene enrichment and functional annotation with KEGG, InterPro, and GO ontologies.

## Data Import

Read both csv files and save them into variables 
```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)

# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```


## DESeq Setup
```{r}
#/ message: false
library(DESeq2)
```

Remove first column of countData so both files (colData and countData) match up

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

Exclude genes that have 0 read count across all samples (columns)
```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[rowSums(countData) != 0, ]
head(countData)
```


## DESeq Analysis

Nice now lets setup the DESeqDataSet object required for the DESeq() function and then run the DESeq pipeline.

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```
```{r}
dds
```

Get results for the HoxA1 knockdown versus control siRNA (labeled as "hoxa1_kd" and "control_sirna" in our original colData metaFile input to DESeq, you can check this above and by running resultsNames(dds) command)

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
summary(res)
```


## Result Visualization

Create a Volcano Plot 

```{r}
plot(res$log2FoldChange, -log(res$padj) )
```

Add some color 

```{r}
# Make a color vector for all the genes
mycols <- rep("gray", nrow(res))

# Override by coloring genes with absolute fold change above 2 (red)
mycols[abs(res$log2FoldChange) > 2] <- "red"

# Color genes with adjusted p-value less than 0.01 and absolute fold change above 2 (blue)
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2) 
mycols[inds] <- "blue"
```

Plot with color

```{r}
plot(res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```


## Add Annotation

Load packages 

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)
```


Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```


## Pathway Analysis

Here we are going to use the gage package for pathway analysis

Download all packages if you dont have them, type this into console:
BiocManager::install( c("pathview", "gage", "gageData") )

Load the packages and setup the KEGG data-sets we need
```{r}
# load packages
library(pathview)
library(gage)
library(gageData)

# data-sets
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# examine first 3 pathways
head(kegg.sets.hs, 3)
```


We used the mapIDs() function above to obtain Entrez gene IDs (stored in res$entrez) and we have the fold change results from DESeq2 analysis (stored in res$log2FoldChange)

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Let's run the gage pathway analysis

```{r}
# to get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

Look into gage
```{r}
?gage
```

Look at bject returned by gage

```{r}
attributes(keggres)
# list with 3 elements 
# can also see this in your Environmnet panel/tab window of RStudio or use the R command str(keggres)
```
```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

Now, let's try out the pathview() function from the pathview package to make a pathway plot with our RNA-Seq expression results shown in color.

To begin with lets manually supply a pathway.id (namely the first part of the "hsa04110 Cell cycle") that we could see from the print out above.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")

# This downloads the pathway figure data from KEGG and adds our results to it. Here is the default low resolution raster PNG output from the pathview() call above
```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?
```{r}
## Focus on top 5 upregulated pathways
keggrespathways_down <- rownames(keggres$less)[1:5]

# Extract the 8 character long IDs part of each string
keggresids_down = substr(keggrespathways, start=1, stop=8)
keggresids_down
```
Pass IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids_down, species="hsa")
```

### Gene Ontology

We can also do a similar procedure with gene ontology. Similar to above, go.sets.hs has all GO terms. go.subs.hs is a named list containing indexes for the BP, CC, and MF ontologies. Let’s focus on BP (a.k.a Biological Process) here.

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```
### Reactome Analysis

Let's now conduct over-representation enrichment analysis and pathway-topology analysis with Reactome using the previous list of significant genes generated from our differential expression results above.

First, Using R, output the list of significant genes at the 0.05 level as a plain text file:

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```


## Save Results
```{r}
write.csv(res, file = "keggrespathways.csv", row.names = TRUE)
write.csv(res, file = "keggrespathways_down.csv", row.names = TRUE)
```



