---
title: "Lab 10 Halloween mini project"
author: "Sofia Lanaspa, A17105313"
format: html
---
Download Halloween Candy Data
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

#### Question 1
Q1. How many different candy types are in this data set?
```{r}
nrow(candy)
rownames(candy)
```


#### Question 2
Q2. How many fruity candy types are in the dataset?
```{r}
sum(candy$fruity) 
```
Learn about win percent
```{r}
candy["Twix", ]$winpercent
```


#### Question 3
Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
```{r}
candy["Reese's Miniatures","winpercent"]

max(candy$pricepercent) #price of most expensive candy
```


#### Question 4
Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat","winpercent"]
```

#### Question 5
Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy["Tootsie Roll Snack Bars","winpercent"]
```

Different technique, same job
```{r}
library(dplyr)

candy%>%
  filter(rownames(candy)=="Haribo Happy Cola") %>%
  select(winpercent)
```


Find fruity candy with a winpercent over 50%
```{r}
candy%>%
  filter(winpercent>50,fruity==1)
#more like english, better logic to obtain this value
```

```{r}
top.candy <-candy[candy$winpercent>50]
top.candy[top.candy]
```

Quick insight into dataser some people like using the skimer package and its 'skim()' function
```{r}
# Since we will only use one function from the package, download that single function instead of the entire package with library
skimr::skim(candy)
```

To do PCA, there will be inaccurate results since the winpercent has a much larger range than the range of 0-1 that other columns do, need to SCALE this to fix the issue

#### Question 6
Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
```{r}

```


#### Question 7
Q7. What do you think a zero and one represent for the candy$chocolate column?
```{r}

```


#### Question 8
Q8. Plot a histogram of winpercent values
```{r}

```

#### Question 9
Q9. Is the distribution of winpercent values symmetrical?
```{r}

```

#### Question 10
Q10. Is the center of the distribution above or below 50%?
```{r}

```

#### Question 11
Q11. On average is chocolate candy higher or lower ranked than fruit candy?
```{r}

```

#### Question 12
Q12. Is this difference statistically significant?
```{r}

```





